---
title: "MycoHalo: measuring melanization in fungus–bacterium confrontation plates"
author: "Mitra Ghotbi"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{MycoHalo: measuring melanization in fungus–bacterium confrontation plates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 4.7,
                      dpi = 60, dev = "jpeg", out.width = "100%")
library(MycoHalo)
```

## 1. The problem MycoHalo solves

In our confrontation assay, four *Zymoseptoria tritici* colonies grow at
fixed positions in the four quadrants of a Petri dish. On treatment plates a
bacterium sits in the centre. We want to know whether, and where, the fungus
changes its **melanization** when it interacts with the bacterium.

Measuring this on a *grayscale* image fails for three reasons:

1. **Grey on grey.** Colonies of *Z. tritici* range from almost white
   (yeast-like, unmelanized) through grey to black (DHN-melanized hyphae).
   The bacterial colony is whitish-cream and its diffusion halo is pale
   yellow. In grayscale all of these share the same brightness range, so a
   single threshold cannot separate fungus, bacterium and halo.
2. **Several cultures in one dish.** Four colonies, one bacterium and a
   halo share the same agar. Colonies can touch each other, grow into the
   halo, or throw off satellite micro-colonies, so "the largest bright blob"
   is not a colony.
3. **Dark on dark.** Strongly melanized, wrinkled colonies on dark agar
   have almost the *same mean colour* as the agar. They differ in surface
   *texture* (ridges and valleys), not in average brightness.

MycoHalo solves these with (i) calibrated colour (CIELAB) instead of gray,
(ii) texture, (iii) a model of the agar background and of the lighting, and
(iv) the *known inoculation layout* as a spatial prior for separating the
cultures.

## 2. Photographing plates (this decides your precision)

Image analysis cannot recover information that the photograph does not
contain. The following protocol typically halves plate-to-plate technical
variation.

* **Fixed geometry.** Use a copy stand, and keep the camera height, lens and
  zoom the same for the whole experiment. Shoot the dish without its lid
  (lids cause reflections and condensation) and perpendicular to the lens.
* **Uniform, diffuse light.** Two lights at 45° on either side, or a light
  box. Avoid a single ring light on the lens axis, which puts specular
  highlights on wrinkled colonies. MycoHalo corrects smooth illumination
  gradients (flat-field), but it cannot correct highlights.
* **Manual exposure and white balance.** Switch off auto-exposure and auto
  white balance, or every photo gets a different colour scale. Expose so that
  the brightest bacterial growth stays below about 245 out of 255. Save RAW or
  high-quality JPEG in the sRGB colour space.
* **A grey card in every photo.** Put a small 18 % grey card (L\* ≈ 50) or a
  ColorChecker next to the dish, under the same light. Passing its position
  to `analyze_plate(reference = ...)` removes exposure and white-balance
  differences between photographs, days and cameras. In the simulations this
  cut the error in absolute melanization from ~7 L\* units to < 0.2.
* **Orientation.** Put a pen mark on the dish at 12 o'clock and always
  photograph with the mark at the top, so colony IDs (TL, TR, BR, BL) always
  mean the same inoculation point.
* **Dark matte background** (black velvet or card) under the dish. This makes
  dish detection trivial.
* **Scale.** Measure once, with calipers, the diameter of the dish edge that
  MycoHalo detects (see the dashed circle in the QC figure), or put a ruler in
  the photo and pass `mm_per_px`.

## 3. Quick start

```{r quick}
# A simulated confrontation plate (replace with your file path)
sim <- simulate_plate(seed = 42, facing_darkening = 4)
res <- analyze_plate(sim$image, id = "demo", verbose = FALSE)
res
```

```{r qc, fig.cap = "Quality-control figure. Left: detected dish (dashed), analysed agar disk (solid), expected inoculation points (+) with their search radius (dotted), colony outlines (one colour per colony), bacterium (yellow), halo (orange), satellites and debris (white). Right: pixel classes; each separated colony is filled with the same colour as its outline. Legends are below the panels."}
plot_qc(res)
```

**Always look at the QC figure** before using the numbers. It is also saved
for every plate in batch mode when you give `qc_dir`.

For your own photographs, replace the paths below with the full paths of
your files (this chunk is a template and is not run here; to try the
package on real data right away, use the shipped plates in the next
section):

```{r real, eval = FALSE}
# confrontation plate: 4 Zymoseptoria + central bacterium (default layout)
res <- analyze_plate("/path/to/your/confrontation_plate.jpg",
                     reference = c(0.08, 0.92, 0.03))   # only if a grey card is in the photo
# control plate: same layout, no bacterium
ctrl <- analyze_plate("/path/to/your/control_plate.jpg",
                      layout = plate_layout(centre = "none"))
```

### Real plates shipped with the package

Two of our photographs travel with the package (`inst/extdata/`), so the
whole pipeline can be tried on real data:

```{r real-plates, fig.width = 10, fig.height = 5.6}
treat <- analyze_plate(example_plate("confrontation"), verbose = FALSE)
ctrl  <- analyze_plate(example_plate("control"),
                       layout = plate_layout(centre = "none"), verbose = FALSE)
rbind(cbind(plate = "bacteria", treat$colonies[, c("colony_id", "area_mm2", "MI_mean")]),
      cbind(plate = "control",  ctrl$colonies[,  c("colony_id", "area_mm2", "MI_mean")]))
plot_qc(treat)
plot_melanization_map(treat, mi_range = c(35, 85), main = "Zymoseptoria + bacterium")
plot_melanization_map(ctrl,  mi_range = c(35, 85), main = "Zymoseptoria control")
```

Your own photos go in any folder on your computer. Only example data
belongs inside the package, in `inst/extdata/`, never in a top-level folder.

### Metadata and batch processing

A whole experiment is analysed from a folder of photos plus a small
metadata sheet: one row per photo. `make_metadata()` writes that sheet for
you, and you only fill in what differs between plates. The `treatment`
column decides the layout: `"control"` plates are analysed without a
central bacterium.

```{r metadata, eval = FALSE}
# 1. create the sheet (file names containing "ctrl" become controls)
make_metadata("photos",
              treatment = "bacteria",
              patterns  = c(control = "ctrl|control"),
              extra     = list(bacterium = "isolate_X", zymo_strain = "strain_Y", day = 7))
# -> photos/plate_metadata.csv:
#    file, plate_id, treatment, bacterium, zymo_strain, medium, day, replicate, photo_date, notes

# 2. fix / complete it in Excel or in R, e.g.
meta <- read.csv("photos/plate_metadata.csv")
meta$treatment[meta$file == "2026_05_26_37.JPG"] <- "control"
write.csv(meta, "photos/plate_metadata.csv", row.names = FALSE, na = "")

# 3. analyse everything; the layout follows the treatment column
out <- analyze_plates("photos", metadata = meta, layout = layout_by_treatment,
                      qc_dir = "qc", reference = c(0.08, 0.92, 0.03))
out$colonies <- correct_facing_bias(out$colonies)   # lighting bias from controls

write.csv(out$colonies, "colonies.csv", row.names = FALSE)
compare_melanization(out$colonies, group = "treatment", reference = "control")
```

Several bacteria in one experiment? Put the isolate in `bacterium` and use
it as the group (`compare_melanization(out$colonies, group = "bacterium")`),
with the controls coded as `"none"`. The same columns are available for
day, strain or medium. The script
`system.file("scripts", "analyse_experiment.R", package = "MycoHalo")` does
all of this in four steps.

**Directional light.** A raised colony is always brighter on the side facing
the lamp, and it casts a shadow on the other side. MycoHalo detects and
removes the cast shadows (dark, smooth and agar-coloured). The remaining
brightness asymmetry depends only on where the colony sits in the frame,
so `correct_facing_bias()` subtracts, for each position (TL, TR, BR, BL),
the mean `delta_MI_facing` of the control plates.

### Different layouts

```{r layouts}
plate_layout()                                   # default: TL, TR, BR, BL + bacterium
plate_layout(centre = "none")                    # controls
plate_layout(radius_mm = 20)                     # inoculated 20 mm from the centre
plate_layout(n_colonies = 3, ids = c("A", "B", "C"), start_deg = -90)
```

The expected positions only need to be *approximately* right: each colony is
searched within `search_frac` (default 0.4) of the dish radius around its
expected point, and colonies are matched to positions closest-first.

## 4. How it works

```
photo ─► read & resize ─► detect dish (RANSAC circle, sub-pixel edge) ─► scale (mm/px)
      ─► agar colour (mode) ─► flat-field (log-quadratic, linear RGB) ─► [grey-card calibration]
      ─► objects = colour outliers ∪ texture outliers
      ─► pixel classes: fungus / bacterium / halo / other   (layout-initialised mixture, EM)
      ─► bacterium = compact central component ─► halo = connected diffusion zone
      ─► colony cores (distance-transform maxima) matched to layout ─► seeded watershed
      ─► half-contrast edge placement ─► measurements, profiles, bootstrap CIs, QC
```

### 4.1 Colour space

Pixel values are converted from sRGB to **CIE 1976 L\*a\*b\*** (D65). CIELAB
is approximately *perceptually uniform*: one unit of L\* is the same visible
change in lightness whether a colony is pale or almost black. This makes
L\* differences comparable across the whole melanization range, which is not
true for raw RGB or grey values (they are gamma-encoded and not
perceptually linear). a\* (green–red) and b\* (blue–yellow) carry the
chromatic information that separates achromatic melanized hyphae from the
warm-coloured bacteria and halo.

### 4.2 Background and illumination

The agar colour is found as the *mode* of the colour distribution inside the
dish (coarse 3-D histogram + mean shift). This works even on crowded plates,
because agar pixels cluster tightly in colour space while colonies and halos
spread along gradients.

Light fall-off towards the image edges (vignetting) or oblique lighting
would make a colony look darker simply because of *where* it is on the dish.
Illumination multiplies linear light, so MycoHalo fits, per linear RGB
channel, a quadratic surface to log-intensity of agar pixels (with 3-MAD
trimming of outliers) and divides it out, normalising every pixel to the
illumination at the dish centre. The QC table reports the illumination range
that was corrected (`illumination_range_pct`).

### 4.3 Which pixels are "something"

A pixel is an **object** (not agar) if either

* its colour differs from agar by more than the 99.9 % quantile of the agar
  noise **and** by ΔE > 3 (≈ one just-noticeable difference), or
* its local texture (SD of L\* in a ~0.2 mm window) exceeds agar texture by
  more than 4 robust SDs. This is what finds heavily melanized, wrinkled
  colonies on dark agar.

### 4.4 Fungus, bacterium or halo?

Each object pixel is described by

* the **magnitude** of its colour difference from agar, log ΔE,
* the **direction** of that difference (a unit vector in L\*a\*b\*),
* its **texture**.

The direction is the key. A diffusible pigment at falling concentration moves
the colour along *one* straight line from the agar colour towards the pigment
colour, so the faint fringe of a halo points the same way as its saturated
core. Grey fungal tissue points along the neutral lightness axis at every
level of melanization. Because the direction of a faint pixel is uncertain,
the model widens each pixel's direction likelihood by its own noise
variance, (σ_noise / ΔE)² (a heteroscedastic, measurement-error-aware
mixture). The halo, being a diffusion gradient, gets a magnitude likelihood
that is flat up to its core level.

The mixture is fitted by EM. It is **initialised from the layout**: textured
pixels near the expected fungal positions seed "fungus", the centre seeds
"bacterium", and smooth pixels in the ring around the centre seed "halo".
The classes therefore keep their biological meaning. A halo class whose
colour direction is within 12° of the fungus is dropped, because it only
modelled colony margins.

Posteriors are smoothed in space before the decision, weighted by
confidence. Faint pixels borrow the class of the confident tissue around
them within 0.5 mm, and pixels that fit no class are marked "other" and are
never measured.

For unusual pairs (a pigmented bacterium, a different fungus) you can
instead train a supervised classifier from a few clicked regions:

```{r supervised, eval = FALSE}
p   <- read_plate("plate.jpg") |> detect_plate() |> model_background()
reg <- annotate_plate(p)                       # click fungus, bacteria, halo regions
clf <- train_classifier(extract_training_pixels(p, reg))
res <- analyze_plate("plate.jpg", classifier = clf)
```

### 4.5 Separating the cultures

* **Bacterium:** the compact bacteria-class component nearest the centre.
  Thin bridges into light colonies are cut by an opening of 0.5 mm.
* **Colony cores:** maxima of the Euclidean distance transform of the
  fungal mask. These are the points deepest inside tissue, found by greedy
  non-maximum suppression and matched to the layout positions
  closest-first.
* **Seeded watershed:** from these cores, a marker-controlled watershed on
  the negative distance transform, restricted to fungal tissue, assigns every
  fungal pixel to one colony. Touching colonies are split along their neck,
  and no colony can claim tissue it is not connected to.
* **Satellites:** fungal fragments not connected to any core are counted
  and excluded.
* **Edges:** outlines are re-placed at *half contrast* (the classical 50 %
  criterion for a blurred step edge), which removes the ~1-pixel outward
  bias of smoothed masks.

## 5. What you get

### Colony table (`res$colonies`)

| column | meaning |
|---|---|
| `area_mm2`, `eq_diameter_mm`, `perimeter_mm` | size (scale from dish diameter) |
| `circularity`, `solidity` | 4πA/P² and area / convex-hull area (lobing, sectoring) |
| `L_mean`, `L_median`, `L_sd`, `L_q05…q95` | CIE lightness of the colony interior |
| **`MI_mean`**, `MI_median` | melanization index 100 − L\* (higher = darker) |
| `MI_lo`, `MI_hi` | 95 % spatial block-bootstrap CI of `MI_mean` |
| `gray_imagej` | mean (R+G+B)/3 on 0–255, as ImageJ/Fiji "8-bit" |
| `a_mean`, `b_mean`, `chroma_mean`, `hue_deg` | colour (brown shift of DHN melanin) |
| `dark_fraction` | fraction of pixels with L\* < `dark_L` (if given) |
| `MI_facing`, `MI_away`, **`delta_MI_facing`** | melanization of the half facing the bacterium vs. the opposite half |
| `R_toward_mm`, `R_away_mm`, **`growth_inhibition_pct`** | radius towards / away from the bacterium; PIRG = 100 (R_away − R_toward)/R_away |
| `gap_bacteria_mm`, `gap_halo_mm` | edge-to-edge distance to bacterium and halo (0 = contact) |
| `specular_fraction`, `offset_from_expected_mm` | QC |

On control plates the "interaction target" is the dish centre, so the same
side-specific metrics give the **null distribution**. Always compare
`delta_MI_facing` and `growth_inhibition_pct` on treatment plates against
controls, never against zero: lobed colonies are never perfectly symmetric.

### Edge-to-centre profiles (`res$profiles`)

The growing margin of a *Z. tritici* colony is young and often lighter; older
central tissue is darker. MycoHalo measures melanization in rings defined by
the **distance from the colony edge** (Euclidean distance transform), not
from the centroid, so the age gradient is followed correctly in lobed
colonies.

```{r profiles, fig.height = 3.5}
sim2 <- simulate_plate(seed = 7, edge_lightening = 10)
res2 <- analyze_plate(sim2$image, verbose = FALSE)
if (requireNamespace("ggplot2", quietly = TRUE)) plot_profiles(res2)
```

Neighbouring pixels are strongly correlated (same hypha, JPEG blocks,
in-camera processing), so a naive pixel bootstrap gives uselessly narrow
CIs. The intervals use a **spatial block bootstrap**: 0.5 mm blocks are
resampled as units.

### Melanization landscape

`plot_melanization_map()` shows the whole plate at once. On the left, a 3D
shaded surface in which each colony rises with a height and sepia colour
given by its local melanization index, over a floor that shows the halo
(gold) and the bacterium (cream). On the right, the same MI field is drawn
over the faded photograph. An arrow on each colony points to the bacterium
and is labelled with the side-specific response ΔMI~facing~.

```{r landscape, fig.width = 10, fig.height = 5.6, fig.cap = "Melanization landscape of a simulated confrontation plate in which each colony is 5 L* units darker on the side facing the bacterium."}
sim3 <- simulate_plate(width = 900, height = 1200, seed = 42,
                       facing_darkening = 5, edge_lightening = 8)
res3 <- analyze_plate(sim3$image, id = "simulated plate", verbose = FALSE)
plot_melanization_map(res3)
```

Use `type = "3d"` or `type = "2d"` for single panels, and fix `mi_range`
when you compare several plates, so that colours and heights mean the same
thing in every figure.

### Plate table (`res$plate_summary`)

Scale, dish fit quality, agar colour and noise, illumination correction,
bacterium area and radius, halo area, outer radius (equivalent and 95th
percentile), halo colour (Δb\* = yellowness relative to agar), satellites,
"other" area, over-exposure and all QC warnings.

## 6. Interpreting melanization in *Z. tritici*

*Z. tritici* produces 1,8-dihydroxynaphthalene (DHN) melanin via a
polyketide synthase pathway (e.g. *PKS1*), under transcriptional control
that includes the *Zmr1* regulator. Melanization varies strongly among
strains and with the environment (Lendenmann et al. 2014; Krishnan et al.
2018). In culture it is usually quantified as the mean grey value of colony
photographs. MycoHalo reports that value (`gray_imagej`) for comparability,
but uses L\* as its primary readout because it is calibrated and
perceptually linear.

Some points to keep in mind:

* **Melanization vs. morphology.** Hyphal, wrinkled colonies cast shadows,
  so part of their darkness is topography rather than pigment. Diffuse
  lighting minimises this. Compare treatments at the same growth stage, and
  consider the edge-distance profile, which separates a darker centre from a
  lighter growing margin.
* **Localized vs. global responses.** A bacterium acting through diffusible
  compounds is expected to create gradients: `delta_MI_facing` and the halo
  metrics (`gap_halo_mm`) capture the spatial component, `MI_mean` the global
  one.
* **Absolute vs. relative.** Without a grey card, absolute L\* shifts with
  exposure. Within one photo session with fixed settings, relative
  comparisons are fine. Across sessions, use `reference`.

## 7. Statistics: the plate is the experimental unit

The four colonies on one dish share medium, bacterium, handling and
photograph, so they are *not* independent replicates. Treating them as
independent is pseudoreplication and inflates significance. Use a mixed
model with a random plate effect (or analyse plate means):

```{r stats}
set.seed(1)
d <- data.frame(plate_id = rep(sprintf("p%02d", 1:12), each = 4),
                treatment = rep(c("control", "bacteria"), each = 24))
d$MI_mean <- 55 + 3 * (d$treatment == "bacteria") +
  rep(rnorm(12, 0, 1.5), each = 4) + rnorm(48, 0, 1)
cmp <- compare_melanization(d, response = "MI_mean", group = "treatment",
                            reference = "control")
cmp$method
cmp$coefficients
cmp$icc      # share of variance between plates
```

A useful design rule: with an intra-plate correlation (ICC) of ~0.5, four
colonies per plate carry the information of only ~1.6 independent colonies.
Add plates, not colonies.

## 8. Validation

Every algorithmic step was validated on simulated plates with known truth
(`simulate_plate()`): grey, wrinkled, lobed colonies, a whitish bacterium,
a diffuse yellow halo overlapping the colonies in brightness, satellites,
vignetting, exposure error and sensor noise. The script is in
`system.file("validation", "validate.R", package = "MycoHalo")`.

```{r validation}
v <- read.csv(system.file("validation", "simulation_validation.csv", package = "MycoHalo"))
agg <- do.call(rbind, lapply(split(v, v$scenario), function(s) data.frame(
  scenario = s$scenario[1],
  detected = paste0(sum(s$detected), "/", nrow(s)),
  max_abs_area_err_pct = max(abs(s$area_err_pct), na.rm = TRUE),
  max_abs_MI_err = max(abs(s$MI_err), na.rm = TRUE),
  mean_delta_MI_facing = round(mean(s$dMI_facing, na.rm = TRUE), 2))))
agg[order(agg$max_abs_MI_err), ]
```

In summary, colony area is within ~2.2 % (typically < 1 %) and
melanization within 0.9 L\* units under all realistic conditions. The exception is an uncorrected
exposure error, which is why the grey card matters. A simulated 5-unit
side-specific darkening is recovered as 4–6 units, against at most ±2 under
the null. A missing colony is reported as not detected rather than replaced by
halo tissue.

Simulations cannot capture every property of real photographs. For new
set-ups, also validate against a few colonies outlined by hand, e.g. in
Fiji.

## 9. Troubleshooting

| QC warning / symptom | cause and fix |
|---|---|
| *Dish detection failed* | background not uniform or not contrasting; use a dark matte background or pass `dish = c(x, y, r)` |
| dish circle on the inner wall instead of the rim | fine, but then set `dish_diameter_mm` to the inner diameter (see §2) |
| *Colony not detected* | failed inoculation, or colony far from its expected position; check `plate_layout()` orientation / `radius_mm`, increase `search_frac` |
| colony includes halo or bacterium | train a supervised classifier (§4.4) for this pair |
| *> 2 % specular pixels* | reduce exposure, diffuse the light |
| *strong illumination gradient* | the correction worked, but improve lighting |
| satellites merged into a colony | they touch the colony in the image; unavoidable, but reported via `solidity` |

## References

* Beucher S, Meyer F (1993) The morphological approach to segmentation: the watershed transformation. In: *Mathematical Morphology in Image Processing*, 433–481.
* CIE (2004) *Colorimetry*, 3rd ed. CIE 15:2004.
* Comaniciu D, Meer P (2002) Mean shift: a robust approach toward feature space analysis. *IEEE TPAMI* 24:603–619.
* Dempster AP, Laird NM, Rubin DB (1977) Maximum likelihood from incomplete data via the EM algorithm. *J R Stat Soc B* 39:1–38.
* Felzenszwalb PF, Huttenlocher DP (2012) Distance transforms of sampled functions. *Theory of Computing* 8:415–428.
* Fischler MA, Bolles RC (1981) Random sample consensus. *Commun ACM* 24:381–395.
* Hurlbert SH (1984) Pseudoreplication and the design of ecological field experiments. *Ecol Monogr* 54:187–211.
* Kasa I (1976) A circle fitting procedure and its error analysis. *IEEE Trans Instrum Meas* 25:8–14.
* Krishnan P, Meile L, Plissonneau C, Ma X, Hartmann FE, Croll D, McDonald BA, Sánchez-Vallet A (2018) Transposable element insertions shape gene regulation and melanin production in a fungal pathogen of wheat. *BMC Biology* 16:78. doi:10.1186/s12915-018-0543-2
* Künsch HR (1989) The jackknife and the bootstrap for general stationary observations. *Ann Stat* 17:1217–1241.
* Lendenmann MH, Croll D, Stewart EL, McDonald BA (2014) Quantitative trait locus mapping of melanization in the plant pathogenic fungus *Zymoseptoria tritici*. *G3* 4:2519–2533. PMID 25360032
* Meyer F (1994) Topographic distance and watershed lines. *Signal Processing* 38:113–125.
* Otsu N (1979) A threshold selection method from gray-level histograms. *IEEE Trans Syst Man Cybern* 9:62–66.
* Sharma G (2003) *Digital Color Imaging Handbook*. CRC Press.
* Wells WM (1986) Efficient synthesis of Gaussian filters by cascaded uniform filters. *IEEE TPAMI* 8:234–239.

## Session information

```{r session-info}
sessionInfo()
```
